---
title: "Analyzing Voting Difficulty"
author: "Moonsoo Kim, Ryan Beaty, and David Dorfman"
subtitle: 'Datasci 203: Lab 1 Part 2'
output:
  pdf_document:
    toc: no
    number_sections: yes
  word_document:
    toc: no
urlcolor: blue
---

\newpage
\setcounter{page}{1}

```{r load packages and set options, include=FALSE}
library(tidyverse) 
library(magrittr)
library(knitr)
library(patchwork)
library(moments)
library(ggplot2)

theme_set(theme_bw())

options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo=FALSE, message=FALSE)
```

## Introduction {-}

<!--
# Begin your report with an introduction to motivate the analysis.
# Introduce the topic area and explain why the research question is interesting.
# The introduction must "do work," connecting the general topic to the specific techniques in the report.
-->

The question of whether either US political party faces more difficulty voting has heightened importance in today's politically charged climate. Disparities in voting experiences can signify systemic issues that undermine democratic integrity. 

Utilizing the 2022 Pilot Study from the American National Election Studies (ANES), this analysis aims to explore the voting difficulties encountered by individuals aligned with each party. We then apply statistical tests to quantify the significance of our findings in answer to the question:  \vspace{-1.2em}
 \begin{quote} \textit{Do Democratic voters or Republican voters experience more difficulty voting?} \end{quote} 


## Conceptualization {-} 
1. Who or what is a voter? 

\vspace{-0.8em}   In the United States, a voter is a citizen who is eligible to participate in federal, state, or local elections by casting a ballot. The basic eligibility criteria for voting in federal elections are: U.S. citizenship, at least 18 years old on election day, residency in the state where one is registering to vote, and not being disqualified due to mental incompetence, criminal convictions, or other specific reasons.


2. Who is a "Republican" and who is a "Democrat"? 

\vspace{-0.8em}  In the United States, Republicans and Democrats are members or supporters of the two major political parties. Republicans generally lean conservative, favoring lower taxes, less regulation, and a strong national defense. Democrats typically lean liberal, advocating for social justice, higher social spending, and international diplomacy. These are generalizations, and individuals within each party may have diverse views. 

For purposes of this analysis, we aim to classify voters into one of these two groups. This simplifying assumption means that varying degrees of support including leaning, identifying, or strongly identifying with a party will be classified as a party supporter (Democrat or Republican). 


3. What is difficulty voting? 

\vspace{-0.8em}  Difficulty voting refers to challenges or obstacles individuals may face when attempting to participate in an electoral process. These can vary from logistical issues to systemic barriers, and may include: \vspace{-0.5em}  

- Voter ID Laws: jurisdictions may require specific identification which not all eligible voters have.\vspace{-0.5em}
  
- Accessibility: Physical or digital barriers can make it challenging for disabled individuals to vote.\vspace{-0.5em}  

- Long Wait Times: Insufficient polling places or resources can result in long lines.\vspace{-0.5em}  

- Voter Suppression: Tactics such as gerrymandering aimed to discourage specific groups from voting.\vspace{-0.5em}  

- Registration Barriers: Complicated or inconvenient voter registration processes can deter participation.\vspace{-0.5em}  




## Operationalization {-} 
<!-- Only after you have informed your reader of what these concepts are can you then describe how you are going to _measure_ these concepts. 

- What would be the best **possible** method of measuring this concept? Is this method possible? Why or why not? 


- What is the best **available** method of measuring this concept? Why have you opted to use this measurement instead of other possibilities? Map the concept definitions that you have written down onto the variables that you are going to use. Describe, precisely, how the variables were generated, if they come from survey data, provide the text of the question that the respondent is reacting to, not the variable name. 


- What, if any, changes have you made to the dataset from how it was provided? Why did you make those changes, how much data was affected, and what are the consequences for any estimates that you produce? -->
  
The best **possible** method of measuring difficulty in voting among Republicans and Democrats would be to get complete information on each eligible voter immediately after the election on how difficult they found it to vote, and what party they support. This might be feasible using extreme mass surveillance and mandatory surveying of all eligible voters. 


However, the best **available** method of measuring this concept is to use existing ANES survey data that measures respondent's party alignment along with reported difficulty voting. 

The ANES 2022 Pilot Study is a cross-sectional survey based on a sample of respondents drawn from YouGov, a platform where respondents receive rewards for completing surveys. The YouGov panel is not necessarily representative of the U.S. population, for which correction factors have been included but not factored in to our analysis. 


The ANES data includes the following key variables of interest to answer the question at hand:

**1. caseid:** unique identifier

**2. pid_x:** summary of Party ID  (1 strong democrat, 2 democrat, 3 lean democrat, 4 independent, 5 lean republican, 6 republican, 7 strong republican)

**3. votehard:** How difficult was it for you to vote?  (-1 inapplicable, legitimate skip; 1 Not difficult at all; 2 A little difficult; 3 Moderately difficult; 4 Very difficult; 5 Extremely difficult)

<!-- 

## Visual Design {-}
- Any plots or tables that you include must follow basic principles of visual design.
  - A plot/figure must have a title that is informative.
  - Variables must be labeled in plain language. As an example, `v20002` does not work for a label.
  - A plot should have a good ratio of information to ink / space on the page.  Do not select a large or complicated plot when a simple table conveys the same information directly.
- Do not include any plot (or R output in general) that you do not discuss in your narrative.
- The code that makes your plot/figure should be included in your report `.Rmd` file, but should not be shown in your final report. To accomplish this, you can use an `echo=FALSE` argument in the code chunk that produces the plot/figure.
-->


## Data wrangling {-}
<!-- 
To answer your research question, you will have to clean, tidy, and structure the data (A.K.A. wrangle). 

- The code to wrangle data should be included with your deliverable somehow. If you choose to include it in your report `.Rmd` file, then it should not be shown in the PDF of your final report. To accomplish this, you can use an `echo=FALSE` argument for the code chunk that does the wrangling. 
  - A better practice -- not strictly necessary for this lab -- would be to write a function that loads and cleans *all* of the data that is being used by your team for its reports. This way, a single function can be run (and evaluated by your reader) for all the loading, cleaning, and manipulating. 
- While we do not want to prohibit you from using additional tools for data manipulation, you should be able to complete this lab with no more than the base `stats` library, plus `dplyr` and `ggplot2` for data manipulation and plotting. Other tools within the tidyverse are available to use, but don't feel like you have to search them out. 
- You will learn more by writing your own function than you would searching for a package that does one thing for your report.
-->


```{r load data}
voting <- read.csv('./datasets/raw/anes_pilot_2022_csv_20221214.csv')
```

We begin by reducing the dataset to the above three variables which will look at the relationship between party affiliation and difficulty voting.

```{r reduce dataset to variables of interest}
# reduce voting to the 3 variables noted
voting2 <- data.frame(voting$caseid, voting$votehard, voting$pid_x)
```


Next, we eliminate independents (273 '4' responses). This enables a direct comparison of the two parties.
```{r simplify party id to binary}
# Update pid_x:  1,2,3 = Democrat 5,6,7 = Republican else NA and remove

voting2$party <- with(voting2, ifelse(voting.pid_x < 4, 'Democrat',
                                 ifelse(voting.pid_x > 4, 'Republican', 'na')))

# Remove independents (273 '4' responses)
voting2 <- subset(voting2, party == 'Democrat' | party == 'Republican')
```

Lastly, we remove rows with voting difficulty response of 'inapplicable / legitimate skip' (263 '-1' responses).
```{r}
# Remove rows with inapplicable values (263 '-1' responses)
voting2 <- subset(voting2, voting.votehard > 0)
```

We then summarize the remaining data using two histograms - one for Democrats, and one for Republicans. Additionally, we layer on the mean of each group to act as a simple summary statistic. 

```{r analyze data, fig.height=3.5, fig.width=7}

# Calculate the means
mean_dems <-mean(voting2$voting.votehard[voting2$party == 'Democrat'])
label_text_dems <- paste("Mean:", round(mean_dems, 2))

mean_reps <-mean(voting2$voting.votehard[voting2$party == 'Republican'])
label_text_reps <- paste("Mean:", round(mean_reps, 2))

# Set up the layout to have 1 row and 2 columns
par(mfrow = c(1, 2))

# Create the first histogram
hist(voting2$voting.votehard[voting2$party == 'Democrat'],
     main = "Democrats \nVoting Difficulty",
     xlab = "Reported Voting Difficulty",
     ylab = "Count of Respondents",
     ylim = c(0, 500))
abline(v = mean_dems, col = "red", lwd = 2, lty = 2)
text(x = 3, y = 500, labels = label_text_dems, col = "red")


# Create the second histogram
hist(voting2$voting.votehard[voting2$party == 'Republican'],
     main = "Republicans \nVoting Difficulty ", 
     xlab = "Reported Voting Difficulty",
     ylab = "Count of Respondents",
     ylim = c(0, 500))
abline(v = mean_reps, col = "red", lwd = 2, lty = 2)
text(x = 3, y = 500, labels = label_text_reps, col = "red")

# Reset the layout to default
par(mfrow = c(1, 1))

```

A preliminary review of the histograms and means of the data confirms the two groups share a non-normal distribution with similar variances. 



## Hypothesis testing {-}
<!-- 
To answer your research question, you will have to execute one of the statistical tests from the course.

- You need to argue, from the statistical principles of the course, why the test you are conducting is the *most appropriate* way to answer the research question. 
-->

Due to the ordinal nature of the difficulty ratings in the ANES data, we must use the Wilcoxon Rank-Sum Test as the most appropriate way to answer the research question. 

Additionally, the data compares independent groups, thus an unpaired test is appropriate. 

The null hypothesis for this test is that there is equal probability that the difficulty values for Democrats are greater than Republicans, as the reverse (Republicans greater than Democrats).



<!-- 
- Although you might not do this for a report at your organization, for this class please list every assumption from your test, and evaluate it (assess whether the assumption is a reasonable reflection of the natural process that generated the data).

- If you identify problems with some assumptions for your test, that does not mean that you should abandon the analysis or hide the problem. If these "limitations" exist, please describe them honestly, and provide your interpretation of the consequences for your test.
-->

\underline{Wilcoxon Rank-Sum Test Assumptions:}

1. **Independence:** This assumption assumption is met as one respondent's answer should not influence or inform the difficulty of voting for another. Geographic clustering effects are a potential risk should respondents from the same region experience the same logistical voting challenges. 
 

2. **Continuous or Ordinal Data:** Met, as the difficulty ratings are ordinal values from 1 to 5. 

<!-- 
3. **Random Sampling:** Each sample is drawn randomly from the population. This assumption is met as the survey data for each respondent are randomized. Section *'5. Sampling And Weighting'* of the ANES pilot user guidebook outlines the random sampling methodology, detailing how this requirement is satisfied.

4. **Shape of Distributions:** Assumes that the shape of the distribution of the two groups is the same. Based on the histograms above, we can confirm that the distributions are similar in shape. 
-->



<!--- The code that executes your test *should* be shown in your report, because it makes very clear the specific test that you're conducting.
-->
```{r echo=TRUE}
wilcox.test(voting2$voting.votehard ~ voting2$party, paired = F)
```



## Test, results and interpretation {-}

<!-- 

- While you can choose to display the results of your test in the report, you also *certainly* need to write about these results. This should be accomplished using [inline code chunks](https://bookdown.org/yihui/rmarkdown-cookbook/r-code.html), rather than by hard-coding / hard-writing output into your written report. An example of this is included in `lab_1_example_solution.Rmd`. 
-->

As the p-value is less than 0.05 at ~.000006, it suggests that the probability(x < y) <> probability(x > y). Therefore we reject the null hypothesis that there is no difference in difficulty voting reported by Democrats vs Republicans. 

The mean value is not necessarily descriptive in and of itself with ordinal data. However, given the reported difficulty range of (1 - no difficulty to 5 - very difficult), we can use the mean to describe the average experience relative to 1 (no difficulty). 

Using the histograms and means shown above, we visually and quantitatively confirm the greater number of Democrats reporting difficulty voting.

In conclusion, the data suggests a statistically significant increase in difficulty voting for Democrat vs Republican voters.


## Limitations {-}
<!-- Please discuss whether any statistically significant results that you find are of *practical significance*. There are many ways to do this, but the best will provide your reader enough context to understand any measured differences in a scale appropriate to your variables. Explain the main takeaway of your analysis and how it relates to the broader context you identified in the introduction. 
-->

Based on ANES data field turnout22 (whether someone voted on November 8), we can confirm that a 223 of 1585 respondents did not vote. This confirms that the difficulty voting was captured even for respondents that ultimately experienced such difficulty that they did not vote. 

Removing the non-voters results in a Wilcoxon Rank-Sum test that still rejects the null, with a P-value of ~0.002. For the purposes of this analysis, we chose to include both actual (voters) and perceived (non-voters) difficulty voting. 


## Practical Significance {-}
If one group consistently finds it harder to vote, their voices may be underrepresented in electoral outcomes. Democrats' materially higher mean of 1.26 vs Republicans' 1.12 suggests a disparity in voting difficulty.

As the sample size of this survey is very small relative to the U.S. population at 1,585 respondents of a total population of ~330M [https://www.census.gov/popclock/], we hesitate to draw conclusions on systemic issues that undermine democratic integrity. 

However, the high degree of significance in the Wilcoxon Rank-Sum test suggest that a more expansive study encompassing a larger and more diverse sample may prove informative. 




----------
<!-- 
# Introduction

The American National Election Studies (ANES) conducts surveys of voters in the United States, with a flagship survey occurring immediately before and after each presidential election.  In this part, you will use the ANES data to address a question about voters in the US.  Your team will conduct a statistical analysis and generate a written report in pdf format.

This is an exercise in both statistics and professional communication.  It is important that your techniques are properly executed; equally important is that your writing is clear and organized, and your argument well justified. 

# Data 

Data for the lab should be drawn from the 2020 American National Election Studies (ANES). You can access this data at [https://electionstudies.org](https://electionstudies.org). This is the official site of the ANES, a project that has [been ongoing since](https://en.wikipedia.org/wiki/American_National_Election_Studies) 1948, and federally funded by the National Science Foundation since 1977. 

To access the data, you will need to register for an account, confirm this account, and then login. The data that you need should come from the **2022 Pilot Study**.

While you're at the ANES website, you will also want to download the codebook, because all of the variables are marked as something like, `V200002` -- which isn't very descriptive without the codebook. 

For a glimpse into some of the intricacies that go into the design of this study, take a look at the introduction to the codebook.  

Like many modern surveys, the ANES includes survey weights, which are used to correct for situations in which members of one demographic group are more likely to respond to the survey than members of another demographic group. (The target proportions are ultimately based on US census data). The survey weights make it possible to generalize from the a population that represents people who take the survey to a population that represents the United States as a whole. These weights are beyond the scope of our class and you are not expected to utilize them. You will still be able to learn about a population model, even if applicability to the US population is limited.

# The research question

Use the ANES data to address the following question:

> Do Democratic voters or Republican voters experience more difficulty voting?

# Guidance from political scientists 

Political identification in the US is a complex phenomenon that is the topic of a large academic literature.  See `./background_literature/petrocik_2009.pdf` for some guidance about how stated political identity might not match with revealed political identity at the ballot box. 

As practical guidance: 

1. Is it reasonable to use the vote that someone cast to identify their party preference in this case? What if someone had so difficult a time voting that they did not cast a ballot? 
2. Please treat individuals who "lean" in one direction or another as members of that party. This means that someone who "Leans Democratic" should be classified as as Democrat; and someone who "Leans Republican" should be classified as a Republican.

# Report guidelines 

This section provides some guidance for you as you write your report. In [`rubric.md`](./rubric.md) we provide you with specific statements of how we will evaluate your report.  

# General guidance {-}
- You should knit an .Rmd file to create your pdf report.
- Your report should be no more than 3 pages in standard latex formatting (i.e. `output: pdf_document`)
- You should assume your reader is familiar with statistics, but has no special knowledge of the ANES survey.
- Your report should contain either a plot or a table that advances the argument. 

-->
